```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Python Cheat Sheet  

## Mathematics

* `+` and `-`: addition and subtraction
* `*` and `/`: multiplication and division
* `**`: exponentials

## Comparisons 

* `<` and `>`: greater than, less than
* `<=` and `>=`: greater than or equal to, less than or equal to

## String Methods 

* `.upper()` and `.lower()`: Convert string to upper 
* `.split()`: Split a string on a delimiter into a list
* `.join()`: Convert a list into a string using a delimiter
* `.rstrip()`, `.lstrip()`, and `.strip()`: Remove unwanted characters from the right side, left side, or both sides of a string

## List methods

* `.append()`: Add an element to the end of a list

## `Numpy`  

* `.readtxt()`: Import a file 
* `array.shape`: Return the number of rows and columns of a `numpy` array object 

# Python 

## Data Types

**Integers** are whole numbers. For example:

* `2`
* `-3`
* `0`

**Floats** are numbers with a decimal point. For example: 

* `1.2`
* `-3.0`
* `26/3` (evaluates to `8.666`)

**Strings** are characters enclosed by single `'` or double `"` quotation marks. Any text enclosed by quotes will be treated as a string.

* `"My Grandpa's deck has no pathetic cards"`
* `'85.3'`

Note the second example – `85.3` is a float; `'85.3'` is a string.

**Booleans** have two possible values: `True` and `False`. These can also be expressed as `1` (true) or `0` (false).

## Variables

Variables are assigned using the `=` sign:

```
variable_name = value
```

For example:

* `composer = "buxtehude"`
* `year = 1637`

The variable name can be almost anything. Here are some general rules to consider when naming a variable:

* The name must start with a letter or underscore
* The name can only consist of letters, numbers, or underscores
* Variables are case sensitive (i.e. Python interprets `my_number`, `MY_NUMBER`, and `My_Number` as different variables)
* Python has a set of "reserved words" that cannot be used as variable names. These are words that already have a set meaning in Python, such as `True`, `False`, `for`, and `if`. A full list can be found [**here**](https://www.programiz.com/python-programming/keywords-identifier).

## Math

### Mathematical operations 

A lot of mathematical operations in Python are straightforward. Here are some of the basic operations we can perform:

* `+` and `-`: addition and subtraction
* `*` and `/`: multiplication and division
* `**`: exponentials

We can perform mathematical operations on values directly:

```{r, engine="python"}
print(2 + 3)
```

Or we can operate on variables: 

```{r, engine="python"}
myValue = 4 
print(myValue**2)
```

Python will automatically convert integers to floats when appropriate: 

```{r, engine="python"}
print(3 + 2.2)
```

We can save the output of an expression as a variable:

```{r, engine="python"}
my_product = 2 * 10
print(my_product)
```

And likewise we can perform mathematical operations on variables, if these variables store numeric data:

```{r, engine="python"}
number1 = 7
print(number1 / 2)
```

### Order of operations

Python follows the usual mathematical order of operations. And like in math, we can use parentheses `()` to enforce a specific order.

```{r, engine="python"}
print(2 * (2 + 2))
```

## Comparisons 

In Python, comparisons will always return a Boolean, i.e. either `True` or `False.` We can use the following syntax to compare values:

* `<` and `>`: greater than, less than
* `<=` and `>=`: greater than or equal to, less than or equal to

For example:

```{r, engine="python"}
print(5.3 < 17)
print(4. >= (8/2))
```

* `==` checks if two values are equal. `!=` checks if two values are different.

```{r, engine="python"}
print(2 == (10 - 8))
```

Note that a single equals sign `=` is used to assign values. However, a double equals sign `==` is used to compare values.

* `var1 = 3` sets the variable `var1` to `3`
* `var1 == 3` checks whether the value of `var1` is `3`

## Built-In Functions and Methods 

### Functions

A **function** is a block of code that performs a task. Python comes with a substantial set of pre-written functions. 

***Text on how a function is formulated - name, parentheses, optionally arguments*** 

***Is it worth talking about named arguments or positional arguments. Be introspective about this***

For example, the `print()` function displays ***FINISH ME***

***OTHER Built-In FUNCTIONS***'

### Methods

Every data type that we use in Python (strings, integers, etc.) is associated with a set of functions unique to the data type. These functions are called **methods**.

The syntax for using a method is:

```
<objectName>.<methodName>()
```

#### Example

The string method `.upper()` is used to convert a string into uppercase letters. We can use it either directly on the string itself:

```{r, engine="python"}
"peppa".upper()
```

or by operating on a variable:

```{r, engine="python"}
pig = "peppa"
pig.upper()
```

#### Modifying objects

Using a method may or not modify the object you run it on. For example, the `.upper()` method shows you the uppercase version of a string but **does not** modify the actual object that you are operating on. Observe the following code block:

```{r, engine="python", eval = FALSE}
pig = "peppa"
pig.upper()
print(pig)
```

```
## peppa
```

`pig` has not been updated. To actually save the result of `.upper()`, we have to assign it to a variable: 

```{r, engine="python"}
pig_caps = pig.upper()
print(pig_caps)
```

Some methods **do** modify underlying variables. For example, the list `.append()` method (which we will learn about later) *does* change the list if operates on: 

```{r, engine="python"}
myList = []
print(myList)
myList.append('apple')
print(myList)
```

`myList` is altered by `myList.append('apple')`. There's no need to type something like `myList = myList.append('apple')` - this would in fact be an error. Whether methods modify the objects they operate on is something you'll have to keep track of on a case-by-case basis as you learn new methods. 

## Type Conversion


## String Methods

### `.upper()` and `.lower()`

The `.upper()` and `.lower()` methods take a string and convert it to uppercase and lowercase, respectively.

```{r, engine='python'}
print("out on the wily, windy moors".upper())
```

```{r, engine='python'}
aria = "Piangerò La Sorte Mia"
print(aria.lower())
```

### .split()

The `.split()` method takes a string and splits it into a list, dividing the list on a **delimiter** (i.e., separator). The delimiter is provided as an argument:

```{r, engine='python'}
print("Newt eye, frog toe".split(','))
```

If no argument is provided, then the string is split on whitespace (that is, it is split whenever a space or tab is encountered).

```{r, engine='python'}
print("Eye of newt    and toe of frog".split())
```

### .join()

The `.join()` method is the inverse of `.split()`: converts a list into a string, with list elements separated by a delimiter. The general syntax is:

```
"<delimiter>".join(<list>)
```

For example:

```{r, engine='python'}
" ".join(["I", "found", "a", "fox", "caught", "by", "dogs"])
```

If we do not provide a delimiter, then the strings are directly concatenated: 

```{r, engine='python'}
"".join(["I", "found", "a", "fox", "caught", "by", "dogs"])
```

### .rstrip(), .lstrip(), .strip()

These three methods remove unwanted characters on the right, left, or both sides of a string. You can provide the characters you want to remove as an argument:

```{r, engine='python'}
"ricercar........,,,,,,".rstrip(",.")
```

Without an argument, the methods remove spaces: 

```{r, engine='python'}
"          ricercar ".lstrip()
```

Note that in the above example we strip the spaces to the left of the main text, but we **do not** remove the spaces from the middle or right end of the text.

## Lists 

Lists allow us to store multiple objects together.

A list is a sequential group of variables, denoted in Python by square brackets `[]`, with individual entries separated by commas. A few of the neat properties of lists are:

* **Ordered**: The list `[1, 5, 3, 7]` will always store those numbers in the same order.
* **Mixed data types**: `["mercury", 13, 5.3, False]` is a valid list which contains every data type we've seen so far.
* **Can contain other lists**: `[[2, 3], "sulfur", 12, 18]`
* **Can contain repeat values**: `["tomato", "tomato", "tomato", "sulfur"]`

### Indexing

What if we want to extract a specific value from a list? We can use **indexing**. To index in Python, we use the following syntax:

```
variable_name[index]
```

where `index` is the number of the item we wish to extract.

```{r, engine="python"}
alchemists = ["Zosimos", "Oresme", "Flamel", "pseudo-Aristotle"]
print(alchemists[1])
```

Notice that when we printed the item at position 1, we printed out the _second_ entry in `alchemists`. This is because in Python, **indexing begins at 0**. To print out the first entry, we would use `alchemists[0]`.

A couple interesting things we can do with indexing:

* To print multiple consecutive items, we can provide two numbers separated by a colon `:`.

```{r, engine="python"}
print(alchemists[0:2])
```

Note that the first number is **inclusive** and the second number is **exclusive**: we include the item at position 0 (`'Zosimos'`), but not the item at position 2 (`'Flamel'`).

* We can index in reverse. To index from the end of a list, we use negative numbers.

```{r, engine="python", eval=FALSE}
print(alchemists[-1])
```
```{r, engine="python", echo=FALSE}
print(alchemists[-1])
```

***
<details> <summary> Nested lists </summary>

How would we extract the number `3` from the list below?

```{r, engine="python"}
my_list = [1, 2, [3, 4], 5]
```

First, we extract the `[3, 4]` list. This is the third item of the outer list, so it is at position 2 (remember, indexing in Python starts at 0). So we can access the interior list with `my_list[2]`:

```{r, engine="python"}
print(my_list[2])
```

Within the interior list, `3` is the first item, so it is at position 0. It can be accessed with the syntax `my_list[2][0]`:

```{r, engine="python", eval=FALSE}
print(my_list[2][0])
```
```{r, engine="python", echo=FALSE}
print(my_list[2][0])
```

</details>
***

#### Indexing Strings

We can also apply indexing to extract substrings from within a string. This is done identically to how we index a list:

```{r, engine='python'}
creature = 'stingray'
print(creature[0:5])
```

### Adding to lists

Finally, we can add entries to the end of a list. We do this with the `append()` method, which is used with the following syntax:

```
list_name.append(item)
```

For example:

```{r, engine="python"}
florilegium = ['marigold', 'thistle', 'wormwood']
florilegium.append('tansy')
print(florilegium)
```

## For Loops 

In Python, we will often want to perform an action more than once. For example, if we have a list, we might want to do the same operatation on every item within the list.

One way to do this is to use a **for loop**, which is structured like this:

```
for <temporary_variable> in <thing to loop through>:
    {do something}
```

For example:

```{r, engine='python'}
stations = ["Oddity", "Londres", "Cherry", "Swedish"]

for i in stations:
  print(i)
```

### For loop walkthrough

Here is how the loop works:

```
for i in stations:
```

In this line, we:

* Define the temporary variable `i`. The name of this variable is arbitrary.
* We say that we are looping through `stations`.

In the first iteration of the for loop, `i` takes on the value of the first item in `stations` (`"Oddity"`). Now we perform all of the indented code, which here is just a `print()` statement.

In the second iteration of the for loop, we set the value of `i` to the second entry in `stations` (`"Londres"`). We execute all of the indented code, printing out `"Londres"`.

We continue on doing this until there is nothing left in `stations`.

In this example, there was a single line in the body of the loop, but a for loop can be arbitrarily long. Here is a longer example:

```{r, engine='python'}
my_list = [1, 4, 6, 9, 10, 2]

for i in my_list:
  i = i + 3
  i = i ** 2
  print(i)
```

### Repeating an action `n` times

We can also use a for loop to perform an action a set number of times, even when we don’t have a list to loop through. To do so, we can use the `range()` function. We'll use this function a ton throughout the course.

`range(<n>)`, where `n` is a single integer, generates a sequence of numbers from `0` to `n` (not including `n` itself). So to run a function 3 times, we would provide the for loop with `range(3)` (which generates the list `[0, 1, 2]`).

```{r, engine='python'}
for i in range(3):
  print(i)
```

So far, the body of our for loop has always referenced the temporary variable `i`. We can also use a for loop to run a block of code repeatedly without actually manipulating `i`:

```{r, engine='python'}
for i in range(3):
  print("All work and no play")
```

## Reading in Data and Text Parsing




***FILL***



## Modules 

Often, we need to make use of functions beyond the basic ones in Python. To do this, we can import a **module**, or a collection of pre-written functions. A module is imported with the following syntax:

```
import <module name>
```

For example, to import the popular plotting module Matplotlib, we write:

```{r, engine='python', eval=FALSE}
import matplotlib.pyplot
```

### Module functions

To use a function from Matplotlib, we need to reference both the package name and the function name, with the general syntax:

```
moduleName.functionName()
```

To use Matplotlib's `show()` function, we would write:

```{r, engine='python', eval=FALSE}
matplotlib.pyplot.show()
```

### Abbreviating module names

What if you don't want to write out `matplotlib.pyplot` every time you run a Matplotlib function? To simplify this, we can give our modules a shorthand name. For example:

```{r, engine='python', eval=FALSE}
import matplotlib.pyplot as plt
```

Now, instead of writing out `matplotlib.pyplot`, we can just write `plt`. The previous `matplotlib.pyplot.show()` command is shortened to:

```{r, engine='python', eval=FALSE}
plt.show()
```

## Numpy 

Numpy is a Python library used for manipulating arrays and performing mathematical operations on matrices. For concision, we will import the `numpy` module with the name `np`, as such:

```{r, engine='python', eval=FALSE}
import numpy as np
```

### Reading in data

The function `np.readtxt()` is used to read in text data. The most basic way to run `np.readtxt()` is:

```{r, engine='python', eval=FALSE}
np.readtxt(<fname>)
```

Consider a hypothetical comma-delimited file `numbers.csv`. The most basic way of reading it in is to run `np.readtxt('numbers.csv')`. However, when we read in data, we typically want to store it in memory for further manipulation, so we usually use `np.readtxt()` in conjunction with variable assignment: 

```{r, engine='python', eval = FALSE}
myNumbers = np.readtxt('numbers.csv')
```


This creates an array where each contains the full text of the line. For example, a component line could appear as: **array(['1,2,3'], dtype='<U5')**, with all variables appearing together as one string. To tell Python that our data is delimited by a comma, we can use the **optional argument**, `delimiter`: 

```{r, engine='python', eval=FALSE}
myNumbers = np.readtxt('numbers.csv', delimiter=',')
```

Now, each number is encoded as its own element:

`array([1, 2, 3])`

Notice that unlike the file name, we need to specify the name of this optional argument (`delimiter=`). Because the file name argument is mandatory, `numpy` always expects the first argument to be the filename. However, since there are many possible optional arguments that once can use, it is impossible to infer which optional argument is being referred to by position alone. For optional arguments, we **always need to specify their name**.

`np.readtxt()` has many useful optional arguments - for example `skiplines`, which allows the user to skip the first `n` lines of a file. A full list of the optional arguments to `np.readtxt()` can be found [here](https://numpy.org/doc/stable/reference/generated/numpy.loadtxt.html) 

### Indexing `numpy` arrays

Within a `numpy` array, we often want to look at a specific element or set of elements. To do this, we use indexing. 

First, we can see the total size of an array by looking at the array's `shape` attribute. An **attribute** is a property inherent to a specific data type; it is typically viewed with the following syntax: `print(<variableName>.<attributeName>)`. We can therefore look at the shape of our array as such: `print(myNumbers.shape)`. This returns the number of rows and columns, respectively: `(45, 12)` - 45 rows and 12 columns.

#### Accessing Single Values

To look within these rows and columns for specific values, we use indexing: 

```{r, engine='python', eval=FALSE}
dataName[<rowNumber>, <columnNumber>]
```

For example, to print the value in the fifth row and third column of `myNumbers`, we would write: 

```{r, engine='python', eval=FALSE}
print(myNumbers[4, 2])
```

Remember that in Python, we start counting at `0`.

#### Accessing Multiple Values 

To print multiple consecutive items, we can provide two numbers separated by a colon `:`.

```{r, engine='python', eval=FALSE}
print(myNumbers[2:4, 5:10])
```

Note that the first number is **inclusive** and the second number is **exclusive**: We will print the elements in rows 3 and 4 and not row 5. We will print the the elements in columns 6 through 10, and not in column 11.

# Math in `numpy`

## Plotting 

In this course, we will use the library Matplotlib for plotting. For concision, we will import the `matplotlib.pyplot` module with the name `plt`, as such:

```{r, engine='python', eval=FALSE}
import matplotlib.pyplot as plt
```









































































