```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Python 

## Data Types, Variables, Math

### Types of Data

**Integers** Integers are whole numbers, without a decimal point. For example:

* 2
* -3 
* 0

**Floats** Floats are numbers with a decimal point. For example: 

* 1.2
* -3.0
* 26/3 (This one doesn't actually have a decimal point written in by me, but is a float because the expression evaluates to 8.666)

**Strings** Strings are character data enclosed by single ' or double " quotation marks. Any text enclosed by quotes will be treated as a string.

* "My Grandpa's deck has no pathetic cards"
* '85.3'

Note the second example – `85.3` is a float; `"85.3"` is a string.

**Booleans** A Boolean has two possible values: `True` and `False`. These can be expressed interchangeably as `1` or `0`, respectively.  

### Variables

A variable is assigned using the equals sign, with this general syntax:

`variable_name = value`. 

For example:

* `composer = "buxtehude"` 
* `year = 1637` 

The data name can be almost anything. Here are the rules to consider when naming a variable:

* A variable name must start with a letter or underscore
* A variable name can only consist of letters, numbers, or underscores
* Variables are case sensitive (i.e. Python would interpret my_number, MY_NUMBER, and My_Number as different variables)
* Python has a set of "reserved words" that cannot be used as variable names. These are words that already have a set meaning in Python, such as 'True', 'False', 'for', and 'if'. A full list can be found here: https://www.programiz.com/python-programming/keywords-identifier

### Mathematical Operations 

A lot of mathematical operations in Python are straightforward. Here are some of the basic operations we can perform

* `+` and `-` – addition and subtraction
* `*` and `/` – multiplication and division
* `**` – Exponentials

We can perform mathematical operations on values directly:

```{r, engine="python"}
print(2 + 3)
```

or we can operate on variables: 

```{r, engine="python"}
myValue = 4 
print(myValue**2)
```

Mathematical conversions will automatically convert integers to floats when appropriate: 

```{r, engine="python"}
print(3 + 2.2)
```

We can save the output of an expression as a variable:

```{r, engine="python"}
my_product = 2 * 10
print(my_product)
```

And likewise we can perform mathematical operations on variables, provided these variables store numeric data: 

```{r, engine="python"}
number1 = 7
print(number1 / 2)
```

## Built-In Functions 

A **function** is a block of code that performs a task. Python comes with a substantial set of pre-written functions. 

***Text on how a function is formulated - name, parentheses, optionally arguments*** 

***Is it worth talking about named arguments or positional arguments. Be introspective about this***

For example, the `print()` function displays ***FINISH ME***

***OTHER Built-In FUNCTIONS***

## Modules 

Often, we need to make use of functions beyond those that are packaged with Python. To do this, we can import a **module**, or a collection of pre-written functions. A module is imported with the following syntax: `import <module name>`. For example, to import the popular plotting module Matplotlib, we write:

```{r, engine='python', eval=FALSE}
import matplotlib.pyplot
```

To use a function from this module, we need to reference both the package name and the function name, with the general syntax: `moduleName.functionName()`. To user Matplotlib's `show()` function, we would write:

```{r, engine='python', eval=FALSE}
matplotlib.pyplot.show()
```

What if you don't want to write out `matplotlib.pyplot` in full each time that you run a Matplotlib function? To simplify this, we can give our modules a shorthand name. For example: 

```{r, engine='python', eval=FALSE}
import matplotlib.pyplot as plt
```

Now, instead of writing out `matplotlib.pyplot`, we can just write `plt`. The previous `matplotlib.pyplot.show()` command is shortened to:

```{r, engine='python', eval=FALSE}
plt.show()
```

### Numpy 

Numpy is a library designed for mathematical operations on matrices and arrays, and is particularly useful for working with tabular data. 

```{r, engine='python}
import numpy
```





















