```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Python 

## Built-In Functions and Methods 

### Functions

A **function** is a block of code that performs a task. Python comes with a substantial set of pre-written functions. 

***Text on how a function is formulated - name, parentheses, optionally arguments*** 

***Is it worth talking about named arguments or positional arguments. Be introspective about this***

For example, the `print()` function displays ***FINISH ME***

***OTHER Built-In FUNCTIONS***'

### Methods

Every data type that we use in Python (strings, integers, etc.) is associated with a set of functions unique to the data type. These functions are called **methods**.

The syntax for using a method is:

```
<objectName>.<methodName>()
```

#### Example

The string method `.upper()` is used to convert a string into uppercase letters. We can use it either directly on the string itself:

```{r, engine="python"}
"peppa".upper()
```

or by operating on a variable:

```{r, engine="python"}
pig = "peppa"
pig.upper()
```

#### Modifying objects

Using a method may or not modify the object you run it on. For example, the `.upper()` method shows you the uppercase version of a string but **does not** modify the actual object that you are operating on. Observe the following code block:

```{r, engine="python", eval = FALSE}
pig = "peppa"
pig.upper()
print(pig)
```

```
## peppa
```

`pig` has not been updated. To actually save the result of `.upper()`, we have to assign it to a variable: 

```{r, engine="python"}
pig_caps = pig.upper()
print(pig_caps)
```

Some methods **do** modify underlying variables. For example, the list `.append()` method (which we will learn about later) *does* change the list if operates on: 

```{r, engine="python"}
myList = []
print(myList)
myList.append('apple')
print(myList)
```

`myList` is altered by `myList.append('apple')`. There's no need to type something like `myList = myList.append('apple')` - this would in fact be an error. Whether methods modify the objects they operate on is something you'll have to keep track of on a case-by-case basis as you learn new methods. 