```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Python 

## Lists 

Lists allow us to store multiple objects together.

A list is a sequential group of variables, denoted in Python by square brackets `[]`, with individual entries separated by commas. A few of the neat properties of lists are:

* **Ordered**: The list `[1, 5, 3, 7]` will always store those numbers in the same order.
* **Mixed data types**: `["mercury", 13, 5.3, False]` is a valid list which contains every data type we've seen so far.
* **Can contain other lists**: `[[2, 3], "sulfur", 12, 18]`
* **Can contain repeat values**: `["tomato", "tomato", "tomato", "sulfur"]`

### Indexing

What if we want to extract a specific value from a list? We can use **indexing**. To index in Python, we use the following syntax:

```
variable_name[index]
```

where `index` is the number of the item we wish to extract.

```{r, engine="python"}
alchemists = ["Zosimos", "Oresme", "Flamel", "pseudo-Aristotle"]
print(alchemists[1])
```

Notice that when we printed the item at position 1, we printed out the _second_ entry in `alchemists`. This is because in Python, **indexing begins at 0**. To print out the first entry, we would use `alchemists[0]`.

A couple interesting things we can do with indexing:

* To print multiple consecutive items, we can provide two numbers separated by a colon `:`.

```{r, engine="python"}
print(alchemists[0:2])
```

Note that the first number is **inclusive** and the second number is **exclusive**: we include the item at position 0 (`'Zosimos'`), but not the item at position 2 (`'Flamel'`).

* We can index in reverse. To index from the end of a list, we use negative numbers.

```{r, engine="python", eval=FALSE}
print(alchemists[-1])
```
```{r, engine="python", echo=FALSE}
print(alchemists[-1])
```

***
<details> <summary> Nested lists </summary>

How would we extract the number `3` from the list below?

```{r, engine="python"}
my_list = [1, 2, [3, 4], 5]
```

First, we extract the `[3, 4]` list. This is the third item of the outer list, so it is at position 2 (remember, indexing in Python starts at 0). So we can access the interior list with `my_list[2]`:

```{r, engine="python"}
print(my_list[2])
```

Within the interior list, `3` is the first item, so it is at position 0. It can be accessed with the syntax `my_list[2][0]`:

```{r, engine="python", eval=FALSE}
print(my_list[2][0])
```
```{r, engine="python", echo=FALSE}
print(my_list[2][0])
```

</details>
***

#### Indexing Strings

We can also apply indexing to extract substrings from within a string. This is done identically to how we index a list:

```{r, engine='python'}
creature = 'stingray'
print(creature[0:5])
```

### Adding to lists

Finally, we can add entries to the end of a list. We do this with the `append()` method, which is used with the following syntax:

```
list_name.append(item)
```

For example:

```{r, engine="python"}
florilegium = ['marigold', 'thistle', 'wormwood']
florilegium.append('tansy')
print(florilegium)
```