```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Python 

## Math

### Mathematical operations 

A lot of mathematical operations in Python are straightforward. Here are some of the basic operations we can perform:

* `+` and `-`: addition and subtraction
* `*` and `/`: multiplication and division
* `**`: exponentials

We can perform mathematical operations on values directly:

```{r, engine="python"}
print(2 + 3)
```

Or we can operate on variables: 

```{r, engine="python"}
myValue = 4 
print(myValue**2)
```

Python will automatically convert integers to floats when appropriate: 

```{r, engine="python"}
print(3 + 2.2)
```

We can save the output of an expression as a variable:

```{r, engine="python"}
my_product = 2 * 10
print(my_product)
```

And likewise we can perform mathematical operations on variables, if these variables store numeric data:

```{r, engine="python"}
number1 = 7
print(number1 / 2)
```

### Order of operations

Python follows the usual mathematical order of operations. And like in math, we can use parentheses `()` to enforce a specific order.

```{r, engine="python"}
print(2 * (2 + 2))
```