```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Python 

## Modules 

Often, we need to make use of functions beyond the basic ones in Python. To do this, we can import a **module**, or a collection of pre-written functions. A module is imported with the following syntax:

```
import <module name>
```

For example, to import the popular plotting module Matplotlib, we write:

```{r, engine='python', eval=FALSE}
import matplotlib.pyplot
```

### Module functions

To use a function from Matplotlib, we need to reference both the package name and the function name, with the general syntax:

```
moduleName.functionName()
```

To use Matplotlib's `show()` function, we would write:

```{r, engine='python', eval=FALSE}
matplotlib.pyplot.show()
```

### Abbreviating module names

What if you don't want to write out `matplotlib.pyplot` every time you run a Matplotlib function? To simplify this, we can give our modules a shorthand name. For example:

```{r, engine='python', eval=FALSE}
import matplotlib.pyplot as plt
```

Now, instead of writing out `matplotlib.pyplot`, we can just write `plt`. The previous `matplotlib.pyplot.show()` command is shortened to:

```{r, engine='python', eval=FALSE}
plt.show()
```
