```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Python 

## Reading in Data

When working in Python, we often want to import data from an external file or, conversely, write our data to a file. To interact with external data objects, we use the `open()` function. This function takes two arguments: the `name` of the file we want to work with, and the `mode` that we want to interact with this file. 

The common modes we use are: 

* `r` - Read. This allows us to import data from an external file into python.
* `w` - Write. This allows us to write our data to a file. If a file with this name already exists, operating in `w` mode **will overwrite existing file contents**.
* `a` - Append. This allows us to write our data to a file. If a file with this name already exists, operating in `a` mode will **add to existing file contents**.

To open a file with the name `insects.txt`, for reading in file contents, we would therefore use the syntax: 

```{r, engine='python'}
f = open('insects.txt', 'r')
```

### Parsing a file

If we try to `print()` f, we get the following output: `<_io.TextIOWrapper name='insects.txt' mode='r' encoding='UTF-8'>`

This is because we have opened the file and saved it as a variable, but we haven't actually read through and manipulated the data which it contains. We have two main ways of doing this in base Python: 

#### 1. Parsing with a `for` loop

One way to look through a file's contents line by line is to use a `for` loop. We can loop through a file with the syntax:

```{r, engine='python'}
for line in f:
  print(line)
```

As it turns out, at the end of each line, there is a special end of line character, `\n`. To just read in the data without the return character, we can use `.strip()`:

```{r, engine='python'}
for line in f:
  print(line.strip('\n'))
```

#### 2. Parsing with `readlines()`

`readlines()` is a method that allows us to read through the entire file all at once, returning file contents as a list:

```{r, engine='python', echo=FALSE}
f = open('insects.txt', 'r')
```

```{r, engine='python', echo=FALSE}
print(f.readlines())
```

This is more concise than a for loop, but all lines are read in without the manipulations that we can perform line-by-line in our loop.

#### Closing files

We can close a file with the `.close()` method. To close `insects.txt` (previously saved as the variable `f`), we would run:

```{r, engine='python', echo=FALSE}
f.close()
```

Once the file is closed, we cannot read it or write to it without opening it again. 