```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Python 

## String Methods

### `.upper()` and `.lower()`

The `.upper()` and `.lower()` methods take a string and convert it to uppercase and lowercase, respectively.

```{r, engine='python'}
print("out on the wily, windy moors".upper())
```

```{r, engine='python'}
aria = "Piangerò La Sorte Mia"
print(aria.lower())
```

### .split()

The `.split()` method takes a string and splits it into a list, dividing the list on a **delimiter** (i.e., separator). The delimiter is provided as an argument:

```{r, engine='python'}
print("Newt eye, frog toe".split(','))
```

If no argument is provided, then the string is split on whitespace (that is, it is split whenever a space or tab is encountered).

```{r, engine='python'}
print("Eye of newt    and toe of frog".split())
```

### .join()

The `.join()` method is the inverse of `.split()`: converts a list into a string, with list elements separated by a delimiter. The general syntax is:

```
"<delimiter>".join(<list>)
```

For example:

```{r, engine='python'}
" ".join(["I", "found", "a", "fox", "caught", "by", "dogs"])
```

If we do not provide a delimiter, then the strings are directly concatenated: 

```{r, engine='python'}
"".join(["I", "found", "a", "fox", "caught", "by", "dogs"])
```

### .rstrip(), .lstrip(), .strip()

These three methods remove unwanted characters on the right, left, or both sides of a string. You can provide the characters you want to remove as an argument:

```{r, engine='python'}
"ricercar........,,,,,,".rstrip(",.")
```

Without an argument, the methods remove spaces: 

```{r, engine='python'}
"          ricercar ".lstrip()
```

Note that in the above example we strip the spaces to the left of the main text, but we **do not** remove the spaces from the middle or right end of the text.
