```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Python 

## Type Conversion

Sometimes, data can be converted from one type to another. 

* `float()` - converts integers and strings (when appropriate) to floats

```{r, engine='python', eval=FALSE}
print(float(3))
print(float("2.57"))
```

```{r, engine='python', eval=FALSE}
3.0
2.57
```

* `int()`converts floats and strings into integers. Note that this function always rounds down if necessary. An exception here: `int()` **does not** allow you to convert a string with a decimal point into an integer - `int("3.62")` will throw an error.  

```{r, engine='python', eval=FALSE}
print(int('3'))
print(int(2.57))
```

```{r, engine='python', eval=FALSE}
3
2
```

Note that for the above functions, it is not always possible to do convert data into a different type. In particular, strings with non-numeric characters cause problems - `int("Dachshund")` will not work. 

* `str()` converts all other data types into strings. 

```{r, engine='python', eval=FALSE}
print(str(3))
print(str(2.57))
print(str(True))
```

```{r, engine='python', eval=FALSE}
'3'
'2.57'
'True'
```

A special case of type conversions comes with `booleans`. Numeric values are **automatically converted** into `booleans`. `0` is equivalent to `False`. **All** non-zero numeric values are equivalent to `True`. Likewise, empty strings (`''`) convert to `False`. All non-empty strings (inlcuding the string `'False'`) convert to `True`.
