# Archaic admixture

In this lab, we'll discuss three common statistics, $D$, $f_{4}$, and the $f_{4}$-ratio, which are used for quantifying and testing hypotheses related to admixture (or "introgression").

#### Learning objectives

After completing this chapter, you'll be able to:

1. Explain how incomplete lineage sorting can create patterns of allele sharing that are discordant with species relationships.
2. Describe how introgression is expected to affect patterns of allele sharing.
3. Interpret the D and f4 statistics and their asssociated p-values.
4. Interpret the f4-ratio ratio statistic and whether it implies differences in archaic admixture between populations.


***


## Setup

### R packages

```{r, results = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(admixr)
```


***


## Data

Get data using `admixr`'s `download_data()` function:

```{r}
# download data into current directory
prefix <- download_data(dirname = ".")
```

```{r}
# read in eigenstrat files
snps <- eigenstrat(prefix)
snps
```


***


## Computing the D statistic

```{r}
?d
```

We give `d()` a list of populations to compute the D statistic for, where:

* `W`: Our population of interest
* `X`: An assumed **unadmixed** population (Yoruba)
* `Y`: The **introgressing** population (Vindija Neanderthal)
* `Z`: Outgroup (chimpanzee)

Compute D between French, Sardinian, Vindija (Neanderthal), and chimpanzee:

```{r}
d_result <- d(data = snps,
              # provide population names to calculate D between
              W = "French", X = "Sardinian", Y = "Vindija", Z = "Chimp")

d_result
```

Convert the Z score into a p-value:

```{r}
d_result <- d_result %>%
  # convert z score into pvalue
  mutate(p = 2 * pnorm(-abs(Zscore)))

d_result
```



***


## Computing D for all populations

```{r}
# create vector of populations of interest
pops <- c("French", "Sardinian", "Han", "Papuan",
          # including three African populations
          "Khomani_San", "Mbuti", "Dinka")

# calculate D
d_result <- d(data = snps,
              W = pops, X = "Yoruba", Y = "Vindija", Z = "Chimp") %>%
  # convert z score into pvalue
  mutate(p = 2 * pnorm(-abs(Zscore)))

d_result
```


***


## Plotting the D statistic

Run the code block to plot the D statistic and standard error for each population.

```{r}
ggplot(d_result,
       aes(fct_reorder(W, D), D, color = p < 0.05)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_errorbar(aes(ymin = D - 1.96 * stderr, ymax = D + 1.96 * stderr), width = 0.5) +
  xlab("Population")
```


***


## f4 statistic

Run the code block below to compute the f4 statistic for all populations.

```{r}
f4_result <- f4(data = snps,
                W = pops, X = "Yoruba", Y = "Vindija", Z = "Chimp") %>%
  # convert z score into pvalue
  mutate(p = 2 * pnorm(-abs(Zscore)))

f4_result
```


***


## f4-ratio statistic

Calculate the f4-ratio using the code block below.

```{r}
f4_ratio_result <- f4ratio(data = snps,
                           X = pops, A = "Altai", B = "Vindija", C = "Yoruba", O = "Chimp") %>%
  # convert z score to pvalue
  mutate(p = 2 * pnorm(-abs(Zscore)))

f4_ratio_result
```


***


## Plotting f4-ratio results

Run the code below to plot the $f_{4}$-ratio values we computed.

```{r}
ggplot(f4_ratio_result,
       aes(fct_reorder(X, alpha), alpha, color = p < 0.05)) +
  geom_point() +
  geom_errorbar(aes(ymin = alpha - 2 * stderr, ymax = alpha + 2 * stderr), width = 0.5) +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(y = "Neanderthal ancestry proportion", x = "Present-day individual")
```


***


## Computing statistics in genomic intervals

```{r}
# get the path to the `regions.bed` file
bed <- file.path(dirname(prefix), "regions.bed")

# option 1: KEEP only these regions for analysis
new_snps_keep <- filter_bed(snps, bed)
# option 2: REMOVE these regions from analysis
new_snps_remove <- filter_bed(snps, bed, remove = TRUE)
```

Re-calculate f4-ratio:

```{r}
# f4-ratio with the regions removed
f4_filtered <- f4ratio(data = new_snps_remove,
                       X = pops, A = "Altai", B = "Vindija", C = "Yoruba", O = "Chimp") %>%
  # convert z score to pvalue
  mutate(p = 2 * pnorm(-abs(Zscore)))

f4_filtered
```

Plot the region-excluded f4-ratios:

```{r}
ggplot(f4_filtered,
       aes(fct_reorder(X, alpha), alpha, color = p < 0.05)) +
  geom_point() +
  geom_errorbar(aes(ymin = alpha - 2 * stderr, ymax = alpha + 2 * stderr), width = 0.5) +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(y = "Neanderthal ancestry proportion", x = "Present-day individual")
```


***


## Homework

#### Learning Objectives

* Practice calculating introgression statistics in `admixr`
* Interpret the biological significance of region-specific values of the $f_4$-ratio

#### Assignment

Follow these steps to create your own genome stratifications for calculating the $f_4$ ratio statistic.

1. Go to the [**UCSC Table Browser**](http://genome.ucsc.edu/cgi-bin/hgTables), where you can find a wide selection of annotations for the human genome.
2. Make sure you set the `assembly:` drop-down box to `Feb. 2009 (GRCh37/hg19)`.
3. Use the `group:` and `track:` menus to select any set of genomic regions.
    * You can click the `data format description` button and scroll to the `Description` section to find out what each annotation represents.
4. Under the `Retrieve and display data` section, set the `output format:` to BED.
5. Enter an `output filename:` (ex: `all_genes.bed`).
6. Click `get output` to download the file.
7. In Posit Cloud, upload your file using the `Upload` button in the `Files` panel (bottom right).
8. Run the code block below to reformat the BED file (this matches the UCSC's chromosome naming format with the format used in the `snps` data):

```{r, eval = FALSE}
# fill in blank with the name of your bed file
system(command = "sed -i 's/chr//g' ________")

# get the path to your bed file
bed <- file.path("________")
```

Compute the $f_4$ ratio statistic within and outside of the genomic intervals. Repeat for another set of genome annotations to contrast Neanderthal ancestry in different genomic elements.