# Simulating evolution

In this lab, we'll build a simulation to explore genetic drift using the Wright-Fisher model.

#### Learning objectives

After completing this chapter, you'll be able to:

1. Describe the phenomenon of genetic drift.
2. Explain why random draws from a binomial distribution are a good way to mimic the effect of drift.
3. Interpret allele frequency patterns that occur as a result of drift.
4. Write a for loop in R.
5. Write a function to run code multiple times with different parameters.


***


## Setup

### R packages

```{r}
library(tidyverse)
```


***


## The `rbinom` function

```{r}
?rbinom
```

`rbinom` takes three arguments:

* `n`: how many times we're drawing from the distribution
* `size`: the size of the population we're sampling from (i.e. `N`)
* `p`: the success probability (i.e. allele frequency)

```{r}
rbinom(n = 1, size = 100, prob = 0.5)
```

In the code block below, change the `rbinom` code so that it outputs the allele frequency (instead of the # of individuals):

```{r}

```


***


## Increasing population size

```{r}
rbinom(n = 1, size = 100, prob = 0.5)
```


***


## Simulating multiple generations

For loop:

```{r}
for (i in 1:3) {
  print(i)
}
```

Modify the for loop to run our `rbinom` command:

```{r}

```

Update AF within the for loop:

```{r}

```

Add a population size variable:

```{r}

```



***


## Plotting AF trajectories

Store AFs in a vector:

```{r}

```

Reformat into a tibble for plotting, adding a column for generation number:

```{r}

```

Plot in the code block below:

```{r}

```



***


## Functions

```{r}
run_sim <- {
  
}
```

Run the function:

```{r}
results <- run_sim(Ne = 1000, freq = 0.5, generations = 10000)

# plot output
ggplot(data = results,
       aes(x = gen, y = afs)) +
  geom_line()
```


***


## Homework

One way to extend our simple Wright-Fisher model is to add in selection as a parameter. Selection affects our model by altering the probability of sampling our allele of interest each generation (e.g., positive selection increases the probability, and negative selection decreases it).

Previously, we assumed that this probability was equivalent to the allele's frequency, or $p = \frac{i}{N_e}$, where $N_e$ is the population size and $i$ is the number of individuals who carry the allele.

Assume that in a model **with** selection, the probability is:

$$
p = \frac{i(1 + s)}{N_e - i + i(1+s)}
$$

where $s$ is the **selection coefficient**, and ranges from `-1` to `1`. Note that in the absence of selection (`s = 0`), this probability is $\frac{i}{N_e}$, or the AF.

#### Learning Objectives

* Practice writing functions in R
* Interpret allele frequency trajectories under selection and drift

#### Assignment

In the code block below, modify your `run_sim` function so that it takes in a selection coefficient `s` as a parameter. Run the simulation a few times with and without (`s = 0`) selection, but keeping other parameters the same (`Ne = 10000`, `freq = 0.5`, `generations = 10000`). What do you notice about the allele frequency trajectories?

Note that most selection coefficients are thought to be _extremely small_: https://journals.plos.org/plosgenetics/article/figure?id=10.1371/journal.pgen.1004697.g002